% =====================================================================
%  Paper 2 -- The Spectral Gap-Statement for Attention Transport Geometry
%  Assembled from the ready Section 6 (Gap-Statement) plus a minimal
%  self-contained preliminaries recap.  NO new theory: the body is the
%  existing Section 6 verbatim; the wrapper only makes it standalone.
% =====================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\graphicspath{{figures/}{./}}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

% The Phi-Transport viewpoint is the conceptual origin of this work but
% is NOT a published reference; this paper is self-contained and does not
% depend on it.  The macro expands to a plain descriptive phrase, no cite.
\newcommand{\PaperOne}{the $\Phi$-Transport viewpoint}

\title{\textbf{The Spectral Gap-Statement:}\\[2pt]
  \large When the Negative Subspace of Attention Transport
  is a Well-Posed Invariant}
\author{Evgeny Vyaltsev \and Daniil Vyaltsev}
\date{\today}

\begin{document}
\maketitle

% =====================================================================
\begin{abstract}
The $\Phi$-Transport framework casts attention as a convex transport
process whose free-energy Hessian governs an instability of mass
transport: a negative eigenvalue produces an exponentially growing
mode. The adaptive rank-local router built on that framework implicitly
assumes that the \emph{number} of such modes, $\dim E_-$, is a
well-defined integer. This paper makes that assumption precise. We show
that $\dim E_-$ is a topological invariant --- threshold-free via the
Riesz spectral projector, and stable under perturbations of operator norm
below half the spectral gap, by the Davis--Kahan theorem --- exactly when
the free-energy Hessian possesses a gap, $\Delta(H)>0$. The three
empirical operating ``windows'' of the framework (energy, separation,
softness) are shown to be three coordinate sections of this single
spectral condition. The link from the invariant to geometric defocusing
requires, in addition, an energy-window restriction on the
Jacobi--Maupertuis curvature, which we state precisely. Finally, we
validate the picture on a trained Llama-3.2-1B: the gap-regime holds for
$96\%$ of attention heads, it is established by an explosive phase
transition at the first layer, and the typical head carries exactly one
dominant transport mode. The geometry the framework predicts is not a
synthetic artefact; it is the typical state of attention in a trained
transformer.
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}

This paper adopts \PaperOne{}: the observation that several
machine-learning primitives --- empirical sorting, radix sorting,
equal-mass occupancy, and LogSumExp attention --- can be cast as special
cases of a single convex transport potential $\Phi$. A consequence
of that viewpoint is an inertial Wasserstein flow whose damped dynamics
are governed by a characteristic polynomial $z^2+\gamma z+\kappa=0$: when
the free-energy Hessian has a negative eigenvalue $\kappa<0$, the flow has
an exponentially growing mode, and transport mass defocuses.

The framework's adaptive rank-local router exploits this: it routes a
query to a low-rank sparse path when the effective number of unstable
transport directions is small. That number is the dimension $\dim E_-$ of
the negative subspace of the free-energy Hessian. The router's empirical
law, $r^\star\approx d_{\mathrm{eff}}-1$, treats $\dim E_-$ as a
well-defined integer --- but a count of negative eigenvalues is only
well-posed if the spectrum actually separates the negative modes from the
rest. When it does not, the count depends on an arbitrary tolerance, and
the router's premise dissolves.

This paper supplies the missing precision. Its single technical object is
the \emph{spectral gap} $\Delta(H)$ of the free-energy Hessian
$H$, and its single thesis is that $\dim E_-$ is a meaningful invariant
exactly inside the regime $\Delta(H)>0$. The contributions are:
\begin{itemize}
  \item A threshold-free definition of $\dim E_-$ via the Riesz spectral
    projector, and a proof (Proposition~\ref{prop:gap:wellposed}) that it
    is well-posed precisely in the gap-regime.
  \item A perturbation-stability theorem
    (Proposition~\ref{prop:gap:dk}): $\dim E_-$ is preserved under any
    symmetric perturbation of operator norm below $\tfrac12\Delta$, by
    Davis--Kahan.
  \item The unification of the framework's three empirical operating
    windows --- energy, separation, softness --- as three coordinate
    sections of the single condition $\Delta(H)>0$.
  \item A precise energy-window caveat
    (Proposition~\ref{prop:gap:defocus}) on the implication
    ``negative Hessian $\Rightarrow$ negative curvature'': it is false
    without an energy restriction, and we give the exact threshold.
  \item An empirical validation on a trained Llama-3.2-1B
    (\S\ref{sec:gap:empirical}): the gap-regime is the typical state of
    attention, established by an explosive phase transition at layer~1.
\end{itemize}

The body of the paper is Section~\ref{sec:gap}. Section~\ref{sec:prelim}
first recaps, self-containedly, the minimal $\Phi$-Transport material the
argument depends on.

% =====================================================================
\section{Preliminaries: the free-energy potential and the inertial flow}
\label{sec:prelim}

This section develops, self-containedly, the three elements of the
$\Phi$-Transport viewpoint that the Gap-Statement builds on: the
free-energy potential, the inertial flow it induces, and the role of
damping. The material is elementary --- the central identity is the
Hessian of the cumulant-generating function of an exponential family ---
and is given here so that the paper depends on no external source.

\paragraph{The free-energy potential.}
\label{sec:prelim:phi}
$\Phi$-Transport represents a family of machine-learning primitives by a
single convex potential. For the case of interest here, the potential is
the free energy
\begin{equation}
  \Phi_\beta(x) \;=\; \frac1\beta\,\log\sum_j e^{-\beta U_j(x)},
  \label{eq:prelim:phi}
\end{equation}
with $\beta$ an inverse temperature and $U_j$ a family of component
potentials. LogSumExp attention is the special case
$U_j(x)=-\langle x,k_j\rangle$, with the keys $k_j$ as component data.

\paragraph{The inertial Wasserstein flow.}
\label{sec:inertial}
The framework's central dynamical result is that hysteresis in the
transport map induces an \emph{inertial} flow in the space of measures:
mass does not follow the gradient of $\Phi_\beta$ directly but obeys a
damped second-order equation,
\begin{equation}
  \ddot x + \gamma\,\dot x \;=\; -\nabla\Phi_\beta(x),
  \label{eq:prelim:inertial}
\end{equation}
with $\gamma$ a damping coefficient. Linearizing about a stationary point
and diagonalizing the Hessian decouples \eqref{eq:prelim:inertial} into
scalar oscillators $\ddot a_k+\gamma\dot a_k+\kappa_k a_k=0$, one per
Hessian eigenvalue $\kappa_k$. The characteristic polynomial
$z^2+\gamma z+\kappa_k=0$ then determines stability: for $\kappa_k<0$ it
has a positive root, and the mode $a_k$ grows exponentially. This is the
transport instability whose \emph{count} the present paper makes precise.

\paragraph{The entropy-induced friction tensor.}
\label{sec:friction}
The damping in \eqref{eq:prelim:inertial} need not be a free scalar: in
the $\Phi$-Transport viewpoint it arises as an entropy-induced friction
tensor, a state-dependent operator whose ellipticity controls whether the
flow is dissipative. For the present paper only the scalar reduction is
needed --- $\gamma$ enters solely through the characteristic polynomial ---
so we use the scalar form throughout; the tensor form is not required for
any result below.

\paragraph{The adaptive rank-local router.}
\label{sec:multiaxis}
Built on the above, the framework's router selects, per query, an
effective rank $r^\star$ for a low-rank sparse attention path. Its
empirical calibration follows the law $r^\star\approx
d_{\mathrm{eff}}-1$, where $d_{\mathrm{eff}}$ is a participation ratio of
the query covariance. The present paper does not modify the router; it
clarifies when the integer $r^\star$ that the router relies on is a
well-defined quantity at all.

\medskip
\noindent
With these four items in hand --- the free-energy potential
\eqref{eq:prelim:phi}, the inertial flow \eqref{eq:prelim:inertial}, the
friction tensor, and the router --- the Gap-Statement of
Section~\ref{sec:gap} is self-contained.

% =====================================================================
%  BODY -- the existing Section 6, included verbatim.
%  (The \section heading inside the file becomes Section 3 here.)
% =====================================================================
\input{section6_body.tex}

% =====================================================================
\section{Related work}
\label{sec:related}

\paragraph{Geometry of the space of measures.}
The picture used here --- a free-energy functional whose Wasserstein
Hessian governs the stability of a transport flow --- sits in the
Otto-calculus tradition. Otto's formal Riemannian structure on the space
of probability measures~\cite{otto2001geometry} and the variational
(JKO) formulation of diffusion~\cite{jordan1998variational} established
that gradient flows of free energy are geodesic-type objects;
Ambrosio, Gigli and Savar\'{e}~\cite{ambrosio2008gradient} and
Villani~\cite{villani2009optimal} give the rigorous metric theory. The
Brenier polar-factorization theorem~\cite{brenier1991polar} is what makes
the static transport map a $W_2$ object. Lott and
Villani~\cite{lott2009ricci} connect displacement convexity to a
synthetic Ricci-curvature bound; the present paper's Jacobi--Maupertuis
curvature and its sign condition are a finite-dimensional, attention-
specific instance of that displacement-curvature circle of ideas. Our
contribution is not the geometry itself but the spectral
\emph{well-posedness} question --- when the negative-eigenvalue count is
an invariant --- which the metric-geometry literature does not address
because it works with the full operator rather than an integer count.

\paragraph{Spectral perturbation theory.}
The two technical pillars are classical. The Riesz projector and the
analyticity of spectral subspaces are from Kato's perturbation
theory~\cite{kato1995perturbation}; the stability of an invariant
subspace under a perturbation small relative to the spectral gap is the
Davis--Kahan $\sin\Theta$ theorem~\cite{davis1970rotation}. The present
paper applies them not to a numerical-linear-algebra problem but to the
free-energy Hessian of attention, where the gap acquires a mechanistic
meaning: it is exactly the condition under which the transport-instability
count is physical.

\paragraph{Attention and efficient transformers.}
Attention~\cite{vaswani2017attention} is treated here as one instance of a
free-energy potential; the LogSumExp form is the bridge. On the systems
side, FlashAttention~\cite{dao2022flashattention} and its decode variants
established the streaming-from-HBM pattern that the $\Phi$-Transport
router builds on. The empirical study is on
Llama-3.2-1B~\cite{grattafiori2024llama} over
WikiText-2~\cite{merity2017pointer}. This paper does not propose a new
attention mechanism; it supplies the spectral foundation under an existing
adaptive router, and asks whether the geometry that foundation predicts is
actually present in a trained model.

\paragraph{Attention as optimal transport.}
The view of attention as a transport process has an established line.
Daneshmand~\cite{daneshmand2024provable} shows that softmax self-attention
layers can simulate gradient descent on the dual of the
entropy-regularized optimal-transport problem, with transformer depth
controlling the approximation accuracy. Litman~\cite{litman2025sdpa}
proves that the scaled-dot-product attention forward pass is the exact
solution of a one-sided entropic optimal-transport problem, and --- most
relevant here --- analyses the \emph{information geometry} that this
formulation induces on the space of attention distributions, via the
Fisher information matrix. The present paper shares the premise that
attention is a transport process, but its object is different. Those works
study the transport \emph{map} (the attention weights themselves) and the
geometry of the \emph{distribution simplex}; we study the Hessian of the
free-energy potential \emph{with respect to the query position}, and the
geometry of the \emph{query configuration space} through the
Jacobi--Maupertuis metric. The Fisher geometry of
Litman~\cite{litman2025sdpa} lives on the output simplex; our
Jacobi--Maupertuis geometry lives on the input domain. They are
complementary descriptions of the same mechanism from opposite ends.

\paragraph{Hessian eigenvalues and spectral gaps in transformers.}
Two neighbouring uses of spectral language must be distinguished from ours
explicitly, since the words coincide while the objects do not. First, the
\emph{loss-function Hessian}: negative eigenvalues of the training-loss
Hessian of deep networks have been studied since
Sagun et al.~\cite{sagun2016eigenvalues}, who identified the
bulk-plus-outliers spectral structure, and Alain et
al.~\cite{alain2019negative}, who examined the role of the negative
eigenvalues directly. That Hessian is taken with respect to the
\emph{model parameters} and describes the optimisation landscape; ours is
taken with respect to the \emph{query position} and describes a transport
landscape at a fixed, trained model --- a different operator on a different
space, and the negative-eigenvalue count means a different thing. Second,
the \emph{attention-matrix spectral gap}: Nait Saada et
al.~\cite{naitsaada2025mindthegap} analyse the gap between the two largest
singular values of the attention matrix itself, and identify it as the
driver of rank collapse. That gap is a property of the row-stochastic
attention matrix; ours is a property of the free-energy Hessian
$H=-\beta\operatorname{Cov}_p(k)$, a symmetric negative-semidefinite
operator. The two are not the same spectral object: theirs separates
signal-propagation modes of a Markov matrix, ours separates the
transport-instability subspace of a curvature operator. We adopt the term
``spectral gap'' in its standard operator-theoretic sense and flag the
coincidence to forestall confusion.

% =====================================================================
\section{Conclusion}
\label{sec:conclusion}

The negative subspace of the free-energy Hessian is the object on which
the $\Phi$-Transport router rests, and this paper has shown when that
object is mathematically well-defined: exactly inside the gap-regime
$\Delta(H)>0$. Inside it, $\dim E_-$ is threshold-free (the Riesz
projector) and perturbation-stable (Davis--Kahan); the framework's three
empirical windows are one spectral condition seen from three coordinate
axes; and the link to geometric defocusing holds under one additional,
precisely-stated energy restriction. The measurement on a trained
Llama-3.2-1B closes the argument: the gap-regime is not a property of
hand-built synthetic geometry but the ordinary state of attention in a
working transformer, reached by an explosive transition at the first
layer, with the typical head carrying exactly one dominant transport
mode. The router's effective rank is therefore a well-posed quantity in
the regime where the router operates --- and the regime $\Delta\to0$, far
from being a numerical nuisance, is the structural signature of the
transition to dense attention.

% =====================================================================
\bibliographystyle{plain}
\bibliography{refs}

\end{document}
